The importance of a complete description of dependence between variables for ecology and related fields

### 2019-05-03 Paper.Rmd
edits

<!--***DAN-->

@@

Across essentially all the sub-fields and levels of organization within ecology for which we 
examined a dataset, copula structures were typically non-normal and variables 
commonly exhibited asymmetric tail 
dependence. To make results easier to absorb for readers who are new to copulas: we first 
present results in detail for an example bivariate dataset; then we summarize results
for all bivariate datasets; then we present results in 
detail for an example multivariate dataset; then we summarize results 
for all multivariate datasets.

<!--bivariate example-->
We present the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}A,E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values also differed substantially 
from each other, answering the third part of Q1. The numbers reflect stronger upper- than lower-tail
dependence for the soil C and N data, and this can also be seen visually in extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both that tail dependence was distinct from that of an N copula for the soil data, and that it 
was asymmetric (Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}, 
Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}).

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes results for Q1 for the bivariate datasets. For all $4$ 
bivariate datasets, the two variables were always significantly related 
(Table \ref{tab_bivar_summary}, row 1). A non-normal copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(Table \ref{tab_bivar_summary}, row 2). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(Table \ref{tab_bivar_summary}, rows 3-4), answering the first part of Q1. 
Lower- and upper-tail statistics (Table \ref{tab_bivar_summary}, rows 7-8) 
differed substantially from 0, the tail dependence statistics of 
the N copula, answering the second part of Q1; and also differed substantially from each other
(Table \ref{tab_bivar_summary}, row 9),
answering the third part of Q1. The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse. 
Although these results convincingly show non-normal copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(Table \ref{tab_bivar_summary}, rows 5-6).
But our nonparametric statistics confirmed the same conclusions 
(Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}) for first three bivariate datasets.
The fourth bivariate datasets we analyzed was above-ground biomass and percent cover data (i.e. 
the data used to compute Shannon's diversity index) for the year 2000, but we did the same analysis 
with Cedar creek data sampled over 168 plots from other available years (1996-1999, 2007) too. For 1996 and 
2007 copula structures showed no dependence between two variables concerned, for other years (1997-1999), 
we found similar results as of year 2000 (i.e. best fit non-Normal copula and overall lower tail-dependence).
Due to smaller sample size relative to other three bivariate datasets, we chose a wider range with 
non-parametric analysis (Appendix \ref{SM-Q1bivarnonparam}). Detailed results for each dataset are in Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} and Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present our analysis of the green spruce aphid abundance data (see Data)
as an example multivariate analysis.
Independence was tested for each pair of the 10 sampling locations, and was 
rejected in every case. 
Model selection among copulas was carried out separately for each pair of locations.
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for the best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than those for the N copula (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1 from the Introduction for these data.
Goodness of fit tests based on Cramer-von Mises or Kolmogorov-Smirnov statistics 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data, demonstrating that the collection of 
copula families we used was sufficiently broad that the copulas of all pairs of 
locations could be characterized by at least one of them.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently substantially large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1 for these data. Apparently these aphid data have greater tail dependence 
(both lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower-tail dependence statistics minus model-averaged 
upper-tail dependence statistics (Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`), 
indicating that lower-tail dependence across space was much stronger than upper-tail 
dependence. Thus the spatial synchrony of rarity in green spruce aphid is stronger 
than the spatial synchrony of outbreak events. This also answers the third part of Q1.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results for 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methane-flux data (see Data). Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
substantially better than AIC weights of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
For these data, best-fitting copulas were nearly always considered an adequate fit, 
according to Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown). The conclusions were generally supportive that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

dt <- tibble(Description = c("1. p-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. p-value, Cramer-von Mises goodness of fit test",
           "6. p-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary[1:9,], 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights.\\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we addressed Q2 by exploring via mathematical 
models and simulations three possible mechanisms of this structure which 
seem likely to operate in diverse ecological and evolutionary scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables in those places that are influenced by the 
environmental variables.

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable such as methane or CO$_2$.

To explore the importance of nonlinear dynamics for our hypothesis, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model strongly into play. 
The $\epsilon_i(t)$ were again standard
normal and independent through time, but were related (see below) across locations.

For each of the three model setups, for each of the copulas C, SC, G, SG, J, SJ,
for each Kendall correlation value $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used the generated noise to drive the model, and retained both the
noise and population values for the final 500 
time steps of the simulation. For each simulation, the following
statistics were computed for both the noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted for noises and populations against $\tau$. If the hypothesis is
correct that copula structure between locations of environmental noise can be 
transmitted to produce similar dependence structure for ecological variables, 
plots should be similar for populations and noises.

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $a=0.2$, and 
$b=0.1, -0.1, 0.5, -0.5$ in different simulations. 
Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
means that the model has a slower return time to equilibrium after a perturbation. 
The $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of the environmental variables.
Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for $25000$ time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final $2500$ time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable, in spite of the absence of such structure in the driver
environmental variables.

The third mechanism we explored is evolutionary, and pertains to bivariate 
trait data across species, e.g., to our bird and mammal BMR versus body mass 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million different
bivariate draws from distributions showing one of five distinct types of copula 
structures: 1) extreme left-tail dependence, 2) moderate left-tail dependence,
3) symmetric tail dependence, 4) moderate right-tail dependence, and 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech} for details). 
All marginal distributions were standard normal, and distributions all had Spearman 
correlation $0.875$ between their two components, so our simulations assess 
the impact only of copula structure in evolutionary character
changes on evolutionary outcomes. For each of the five copula structures, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across the phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

The hypothesis that copula structure and tail dependence between environmental 
variables measured through time in different locations can produce similar 
structure in ecological variables influenced by the environmental variables was
largely confirmed, though similarities between environmental- and 
ecological-variable copula structure and tail dependencies can be reduced when the
intrinsic dynamics of the ecological variable are strongly nonlinear.
For simulations using AR(1) models (see section on methods for Q2), copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.

>>>>

@@

Across essentially all our datasets, copula structures were typically non-normal and 
commonly showed asymmetric tail 
dependence. To make results easier to absorb: we first 
present results in for an example bivariate dataset; then we summarize results
for all bivariate datasets; then we present results for an example multivariate dataset; then we summarize for all multivariate datasets.

<!--bivariate example-->
For the soil C and N data (Data, Table \ref{tab_data_info}, 
Fig. \ref{fig_biv_multi_raw_cop}A,E), variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent.
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted our 16 copulas
(previous section) to the data and computed AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions (Table \ref{SM-tab_soilCNfit}).
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
Model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1. The values also differed substantially 
from each other, answering the third part of Q1. The numbers reflect stronger upper- than lower-tail
dependence, and this is also visible in the extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

Although these results convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest`). However, nonparametric statistics verified
both that tail dependence was distinct from that of an N copula, and that it 
was asymmetric (Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}, 
Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}).

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes Q1 results for the bivariate datasets. The two 
variables were significantly related for all datasets 
(Table \ref{tab_bivar_summary}, row 1). A non-N copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(Table \ref{tab_bivar_summary}, row 2). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(Table \ref{tab_bivar_summary}, rows 3-4), answering the first part of Q1. 
Lower- and upper-tail dependence statistics often differed substantially from 0 
(Table \ref{tab_bivar_summary}, rows 7-8) and from each other (Table \ref{tab_bivar_summary}, row 9), 
answering the second and third parts of Q1. The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse, though weakly. 
Although these results convincingly show non-normal copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(Table \ref{tab_bivar_summary}, rows 5-6).
But our nonparametric statistics confirmed the same conclusions 
(Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}).
We did the same analysis 
with Cedar Creek data from other available years, 1996-1999 and 2007. For 1996 and 
2007, biomass and Shannon's index were independent; for 1997-1999, results 
were similar to 2000: the best-fitting copula was not N, and there was slightly stronger lower- 
than upper-tail dependence. Detailed results for each dataset are in 
Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} and 
Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present green spruce aphid abundance results (Table \ref{tab_data_info})
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-N
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than N-copula weights (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1. Goodness of fit tests 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data; the collection of 
copula families we used was apparently sufficiently broad that all pairs of 
locations could be characterized by at least one of them.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1. These data have greater tail dependence 
(lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower- minus upper-tail dependence statistics 
(Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of Q1.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--***DAN: You could maybe move the next para to sup mat, though that would complicate the
presentation of skewness results below. Only do it if you are desperate to cut.-->
<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes Q1 results from
multivariate data. Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown); conclusions supported that 
non-N copula structure and tail dependence, and asymmetry of tail dependence, are 
common.

@@

dt <- tibble(Description = c("1. p-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. p-value, Cramer-von Mises goodness of fit test",
           "6. p-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary[1:9,], 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. Cedar Creek results are for the year 2000.\\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common, we addressed Q2 by exploring, using 
models, three possible mechanisms. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables, measured through time 
in two places, may beget similarly 
structured dependence between ecological variables measured through time in those places. 
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
We initially assume the AR(1) dynamics 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The $\epsilon_i(t)$ represent environmental noise and are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The variable, $E$, is very general, and could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane or CO$_2$, or other quantities.

To explore the importance of nonlinear dynamics for our proposed mechanism, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the three model setups above, $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted for noises and populations against $\tau$. If the hypothesis is
correct that copula structure in environmental noises begets similar dependence 
structure for ecological variables, 
plots should be similar for populations and noises.

Our second mechanism, also summarized in the 
Introduction,
relates to Liebig's law of the minimum and to nonlinear 
influences of environmental variables on ecological variables. 
If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored whether the ecological variable
could then exhibit asymmetric tail dependence across space. 
Let $E_i(t)$ again be an ecological variable, $i=1,2$. Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran-like 
mechanism discussed previously does not operate here.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our model selection algorithms and 
non-parametric statistics (methods for Q1) to these 
outputs to discover if the model could
produce non-N copula structure and asymmetric tail dependence in the 
$E_i(t)$.

Our third mechanism is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme left-tail dependence, 2) moderate left-tail dependence,
3) symmetric tail dependence, 4) moderate right-tail dependence, and 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copula structures, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

The hypothesis that copula structure in ecological variables can arise through
Moran-like effects was largely confirmed, though similarities between environmental- and 
ecological-variable copula structure can be reduced when dynamics are strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
